Holomorphic field theories and higher algebra

### 2025-01-28 draft.tex
modest edits plus refs

The mathematical motivation is simple:
for most classical holomorphic field theories, the solutions to the equations of motion form a moduli space of natural interest to complex geometers,
and one might hope that the quantization of this moduli space is equally interesting.
A well-known example is holomorphic Chern-Simons theory,
which lives on a Calabi-Yau 3-fold~$X$ and involves the moduli space $\RR\Bun_G(X)$ of holomorphic principal $G$-bundles on~$X$.
(We add the prefix $\RR$ because it is best to use the {\em derived} moduli space.
We do not assume the reader is versed in derived geometry and will offer motivations, glosses, and references throughout the survey.)

@@

In physics, a classical field theory is usually described in terms of a few pieces of data:
\begin{itemize}
\item a manifold $M$ that plays the role of ``spacetime,''
\item a space of ``fields,'' typically the sections of some fiber bundle $F \to M$ or connections on a principal bundle on $M$, and
\item a system of partial differential equations (PDE) for the fields, known as the ``equations of motion.''
\end{itemize}
These equations of motion are variational: 
the solutions to these PDE are the critical points, in a sense, of some function,
called an ``action functional.''
(Variational PDE are often called Euler-Lagrange equations.)
In practice an action functional is a formal expression, 
written as an integral over $M$ of some polydifferential operator on the fields that takes values in densities of~$M$.

@@

\begin{eg}[Massless scalar field theory]
\label{eg: massless scalar}
Let $M$ be a smooth manifold and equip it with a Riemannian metric~$g$.
Let $\Delta_g$ denote the associated Laplace-Beltrami operator, 
and let $\d{\rm vol}_g$ denote the associated volume form.
In the case $M = \RR^n$ with the Euclidean inner product, the Laplace-Beltrami operator is the differential operator
\[
\Delta = \frac{\partial^2}{\partial x_1^2} + \cdots + \frac{\partial^2}{\partial x_n^2}.
\]
Let the space of fields be the smooth functions $C^\infty(M)$.
A smooth function $\phi$ is called, in physics, a {\em scalar} field.
The action functional
\[
S(\phi) = \int_M \phi \Delta_g \phi \, \d{\rm vol}_g
\]
corresponds to the equation of motion $\Delta_g \phi = 0$. 
Thus this scalar field theory has the moduli space of {\em harmonic functions} as its space of solutions.
\end{eg}

@@

\begin{eg}[Chern-Simons theory]
Pick a compact Lie group $G$ and equip it with an invariant metric,
which induces a $G$-invariant inner product $\kappa$ on its Lie algebra~$\fg$.
(The standard choice is the Killing form.)
Take an oriented smooth 3-manifold $M$, 
and consider the space of connections on the trivial principal $G$-bundle on $M$.
This space of fields is isomorphic to $\Omega^1(M, \fg)$, the $\fg$-valued one-forms, 
by viewing $\d + A$ as a connection where $A \in \Omega^1(M, \fg)$.
The equation of motion is the zero-curvature equation
\[
F_A = 0,
\]
which can be expressed equivalently as the Maurer-Cartan equation
\[
\d A + \frac{1}{2} [A,A] = 0.
\]
These equations arise as the Euler-Lagrange equation for the action functional
\beqn
\label{eqn:CS}
CS(A) = \frac12 \kappa(A, \d A) + \frac16 \kappa(A, [A,A]),
\eeqn
the integral of a Chern-Simons form.
Hence this theory is known as {\em Chern-Simons theory}.\footnote{Note that $M$ needs to be an oriented 3-manifold for this action functional to make sense.
The equation of motion exists for a principal bundle on any manifold of any dimension,
but it is only variational for oriented 3-manifolds.}
These equations do not depend on a metric on $M$ or any other geometric structure,
and hence define a ``topological'' field theory in the sense that they are independent of geometry.
This theory is interesting because the moduli space of solutions consists of the flat $G$-bundles on $M$,
a topic of deep interest in geometry and topology.
For physicists, it is interesting because it does not involve geometry, 
by contrast to earlier examples of field theories,
and yet exhibits rich phenomena (notably applications to the quantum Hall effect and 3-dimensional gravity).
\end{eg}

@@

Derived geometry is a sophisticated subject but the motivations and first steps are surprisingly simple,
as we now attempt to explain.
In this paper we only need the reader to understand the rough outline of derived geometry to follow our discussion,
but there are several, much more substantial expositions now available.
(For algebraic geometers consider \cite{TV, Lurie}; for symplectic or Poisson geometers consider \cite{PTVV, Safronov}; for differential geometers consider \cite{Joyce, Carchedi, Steffens}; and for representation theorists consider \cite{BZNad, GaiRoz, Raskin}.)

Before saying anything technical, we indicate how derived geometry generalizes more familiar topics.
The starting place is the theory of vector spaces, which is very concrete and provides a rich source of both geometric intuitions and algebraic tools.
This linear setting generalizes in two directions:
\begin{itemize}
\item the geometry of nonlinear spaces, whether in the style of algebraic geometry or in the style of differential geometry, and
\item homological algebra, or differential graded vector spaces (and modules, sheaves, and so on).
\end{itemize} 
Both of these generalizations pass back to linear algebra in natural ways:
\begin{itemize}
\item at a point on a manifold or variety, the tangent space offers the best linear approximation to the space at that point (where it makes sense), and
\item given a cochain complex $(V,d)$, any cohomology group $H_k(V)$ is a vector space encoding essential information about the complex.
\end{itemize} 
As a first pass at derived geometry, one might say that it seeks a common generalization of nonlinear geometry and homological algebra.
With slightly more precision, one might ask for a notion of space such that each point has a tangent {\em complex}, i.e., the linearization of a point yields a cochain complex.

@@

The essential idea is very simple.
In algebraic geometry, if we have two schemes mapping to another scheme
\[
\Spec(A) \to \Spec(C) \leftarrow \Spec(B),
\]
then the fiber product, or intersection, is
\[
\Spec(A) \times_{\Spec(C)} \Spec(B) = \Spec(A \otimes_C B).
\]
Derived algebraic geometry says that the derived fiber product should be encoded by the derived tensor product~$A \otimes_C^{\LL} B$ instead.
(This insight was first sharply articulated by Serre~\cite{}.)
Thus the underlying underived intersection is encoded in 
\[
H^0(A \otimes_C^{\LL} B) = A \otimes_C B
\]
but there are sometimes ``derived corrections'' in negative cohomological degrees.

For example, given a {\em non}\/regular sequence of irreducible [????] polynomials $\mathbf{f} = (f_1,\ldots,f_k) \in \CC[x_1,\ldots,x_n]$, 
then the underived intersection of the hypersurfaces $V_j = \{f_j = 0\}$ is some quite singular subvariety $X$ of $\AA^n$.
A resolution of its algebra $\CC[X]$ over the polynomials is very complicated.
By contrast the derived intersection is encoded by the Koszul complex determined by the sequence $\mathbf{f}$,
which is a very simple cochain complex (in fact, a semi-free dg commutative algebra). 
It remembers information about the hypersurfaces (and hence $\mathbf{f}$), whereas the underived intersection only knows about the ideal generated by~$\mathbf{f}$.
In this sense it is a ``smarter'' or less ``forgetful'' limit.
KOSZUL HOMOLOGY?

@@

There are holomorphic versions of all these basic types, 
and we now spell out some simple examples.\footnote{See \cite{DonThom} for a pioneering and illuminating discussion of how to find such analogs. See Nekrasov's thesis \cite{NekThesis} for a master physicist's view on holomorphic field theories, before he  reconfigured our understanding of much of gauge theory.}
To give the readers --- possibly a mixed audience of mathematicians and physicists --- some feel for the examples, 
we describe them in several ways: 
in terms of PDE and action functionals but also as moduli spaces.
We do not offer a thorough justification of the dictionary, for which see \cite{CosSUSY} for a wonderful and motivating exposition,
and see \cite{AlfonsiYoung} and \cite{Steffens23} for recent work on this topic.

>>>>

The mathematical motivation is simple:
for most classical holomorphic field theories, the solutions to the equations of motion form a moduli space of natural interest to complex geometers,
and one might hope that the quantization of this moduli space is equally interesting.
A well-known example is holomorphic Chern-Simons theory,
which lives on a Calabi-Yau 3-fold~$X$ and involves the moduli space $\RR\Bun_G(X)$ of holomorphic principal $G$-bundles on~$X$.
(We add the prefix $\RR$ because it is best to use the {\em derived} moduli space.
We do not assume the reader is versed in derived geometry and will offer motivations, glosses, and references throughout the survey.
See Section~\ref{sec: DAG} for a start.)

@@

In physics, a classical field theory is usually described in terms of a few pieces of data:
\begin{itemize}
\item a manifold $M$ that plays the role of ``spacetime,''
\item a space of ``fields,'' typically the sections of some fiber bundle $F \to M$ or connections on a principal bundle on $M$, and
\item a system of partial differential equations (PDE) for the fields, known as the ``equations of motion.''
\end{itemize}
These equations of motion are variational: 
the solutions to these PDE are the critical points of some function on the space of fields,
called an ``action functional,''
and such PDE are often called Euler-Lagrange equations.
(One must take a little care in interpreting the claim of taking critical points.)
In practice an action functional is a formal expression, 
written as an integral over $M$ of some polydifferential operator on the fields that takes values in densities of~$M$.

@@

\begin{eg}[Massless scalar field theory]
\label{eg: massless scalar}
Let $M$ be a smooth manifold and equip it with a Riemannian metric~$g$.
Let $\Delta_g$ denote the associated Laplace-Beltrami operator, 
and let $\d{\rm vol}_g$ denote the associated volume form.
In the case $M = \RR^n$ with the Euclidean inner product, the Laplace-Beltrami operator is the differential operator
\[
\Delta = \frac{\partial^2}{\partial x_1^2} + \cdots + \frac{\partial^2}{\partial x_n^2}.
\]
Let the space of fields be the smooth functions $C^\infty(M)$.
A smooth function $\phi$ is called, in physics, a {\em scalar} field.
The action functional
\[
S(\phi) = \int_M \phi \Delta_g \phi \, \d{\rm vol}_g
\]
yields the equation of motion $\Delta_g \phi = 0$. 
Thus this scalar field theory has the moduli space of {\em harmonic functions} as its space of solutions.
\end{eg}

@@

\begin{eg}[Chern-Simons theory]
Pick a compact Lie group $G$ and equip it with an invariant metric,
which induces a $G$-invariant inner product $\kappa$ on its Lie algebra~$\fg$.
(The standard choice is the Killing form.)
Take an oriented smooth 3-manifold $M$, 
and consider the space of connections on the trivial principal $G$-bundle on $M$.
This space of fields is isomorphic to $\Omega^1(M, \fg)$, the $\fg$-valued one-forms, 
by viewing $\d + A$ as a connection where $A \in \Omega^1(M, \fg)$.
The equation of motion is the zero-curvature equation
\[
F_A = 0,
\]
which can be expressed equivalently as the Maurer-Cartan equation
\[
\d A + \frac{1}{2} [A,A] = 0.
\]
These equations arise as the Euler-Lagrange equation for the action functional
\[
S(A) = \int_M CS(A)
\]
where $CS$ denotes the Chern-Simons form
\beqn
\label{eqn:CS}
CS(A) = \frac12 \kappa(A, \d A) + \frac16 \kappa(A, [A,A]).
\eeqn
Hence this theory is known as {\em Chern-Simons theory}.\footnote{Note that $M$ needs to be an oriented 3-manifold for this action functional to make sense.
The equation of motion exists for a principal bundle on any manifold of any dimension,
but it is only variational for oriented 3-manifolds.}
These equations do not depend on a metric on $M$ or any other geometric structure,
and hence define a ``topological'' field theory in the sense that they are independent of geometry.
This theory is interesting because the moduli space of solutions consists of the flat $G$-bundles on $M$,
a topic of deep interest in geometry and topology.
For physicists, it is interesting because it does not involve geometry, 
by contrast to earlier examples of field theories,
and yet exhibits rich phenomena (notably applications to the quantum Hall effect and 3-dimensional gravity).
\end{eg}

@@

Derived geometry is a sophisticated subject but the motivations and first steps are surprisingly simple,
as we now attempt to explain.
In this paper we only need the reader to understand the rough outline of derived geometry to follow our discussion,
but there are several, much more substantial expositions now available.
(For algebraic geometers consider \cite{Toen,PridEug}; for symplectic or Poisson geometers consider \cite{PTVV, Safronov}; for differential geometers consider \cite{Joyce}; and for representation theorists consider \cite{BZNad, GaiRoz, Raskin}.)

Before saying anything technical, we indicate how derived geometry generalizes more familiar bits of mathematics.
The starting place is the theory of vector spaces, which is very concrete and provides a rich source of both geometric intuitions and algebraic tools.
This linear setting generalizes in two directions:
\begin{itemize}
\item the geometry of nonlinear spaces, whether in the style of algebraic geometry or in the style of differential geometry, and
\item homological algebra, or differential graded vector spaces (and modules, sheaves, and so on).
\end{itemize} 
Both of these generalizations pass back to linear algebra in natural ways:
\begin{itemize}
\item at a point on a manifold or variety, the tangent space offers the best linear approximation to the space at that point (where it makes sense), and
\item given a cochain complex $(V,d)$, any cohomology group $H_k(V)$ is a vector space encoding essential information about the complex.
\end{itemize} 
As a first pass at derived geometry, one might say that it seeks a common generalization of nonlinear geometry and homological algebra.
With slightly more precision, one might ask for a notion of space such that each point has a tangent {\em complex}, i.e., the linearization of a point yields a cochain complex.

@@

The essential idea is very simple.
In algebraic geometry, if we have two schemes mapping to another scheme
\[
\Spec(A) \to \Spec(C) \leftarrow \Spec(B),
\]
then the fiber product, or intersection, is
\[
\Spec(A) \times_{\Spec(C)} \Spec(B) = \Spec(A \otimes_C B).
\]
Derived algebraic geometry says that the derived fiber product should be encoded by the derived tensor product~$A \otimes_C^{\LL} B$ instead.
Thus the underlying underived intersection is encoded in 
\[
H^0(A \otimes_C^{\LL} B) = A \otimes_C B
\]
but there are sometimes ``derived corrections'' in negative cohomological degrees.

For example, given a {\em non}\/regular sequence of irreducible polynomials $\mathbf{f} = (f_1,\ldots,f_k) \in \CC[x_1,\ldots,x_n]$, 
then the underived intersection of the hypersurfaces $V_j = \{f_j = 0\}$ is some quite singular subvariety $X$ of $\AA^n$.
A resolution of its algebra $\CC[X]$ over the polynomials is very complicated.
By contrast the derived intersection is encoded by the Koszul complex determined by the sequence $\mathbf{f}$,
which is a very simple cochain complex (in fact, a semi-free dg commutative algebra). 
It remembers information about the hypersurfaces (and hence $\mathbf{f}$), whereas the underived intersection only knows about the ideal generated by~$\mathbf{f}$.
In this sense it is a ``smarter'' or less ``forgetful'' limit.

@@

There are holomorphic versions of all these basic types, 
and we now spell out some simple examples.\footnote{See \cite{DonThom} for a pioneering and illuminating discussion of how to find such analogs, by mathematicians. See Nekrasov's thesis \cite{NekThesis} for a master physicist's view on holomorphic field theories, before he  reconfigured our understanding of much of gauge theory.}
To give the readers --- possibly a mixed audience of mathematicians and physicists --- some feel for the examples, 
we describe them in several ways: 
in terms of PDE and action functionals but also as moduli spaces.
We do not offer a thorough justification of the dictionary, for which see \cite{CosSUSY} for a wonderful and motivating exposition,
and see \cite{AlfonsiYoung} and \cite{Steffens23} for recent work on this topic.
